Mike Ciaccio
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(plyr)
library(GGally)
library(ggpubr)
library(RColorBrewer)
library(scales)
library(dplyr) 
library(gridExtra)
library(grid)
library(tidyverse)
```
 
# Univariate Plots Section
 
This report investigates the wineQualityReds.csv dataset, consisting of  
13 variables for 1599 observations.  

Quick look - column names and first observation  
```{r echo=FALSE, Load_the_data}
# Load the Data
# Instantiate R data frame from csv file.
wineQualityReds = read.csv("wineQualityReds.csv")
```

```{r echo=FALSE, Echo_dataset_head}
head(wineQualityReds, 1)
```
<br>
structure - wineQualityReds.csv  
```{r echo=FALSE, Echo_dataset_str}
# Echo dataset structure - wineQualityReds.csv
str(wineQualityReds)
```
<br>

##### Initial Exploratory Data Analysis  
My primary interest is the relationship between red wine quality and  
alcohol content. Also of interest is the relationship between pH,  
and the other 3 acidity metrics, fixed.acidity, volatile.acidity, citric.acid.  

Drill down into the individual attributes.  
Red Wine quality analysis.
```{r echo=FALSE, warning=FALSE,message=FALSE, Echo_dataset_quality, comment = NA, eval=TRUE}
summary(wineQualityReds$quality)
ggplot(data=wineQualityReds, aes(x=quality)) + geom_bar(color = 'black', 
                                                        fill = '#ADD8E6') +
labs(title = 'Red Wine quality visualization', x = "Quality rating", y = 'Count') 
```
<br>
The quality attribute approximates a normal distribution.
<br>
<br>

Red Wine alcohol analysis.
```{r echo=FALSE, warning=FALSE,message=FALSE, Echo_dataset_alcohol, comment = NA, eval=TRUE}
summary(wineQualityReds$alcohol)

ggplot(aes(x=alcohol), data=wineQualityReds) + 
  geom_histogram(color = 'black', fill = '#ADD8E6', 
                 stat = 'bin', binwidth = 0.3) + 
  xlab("Alcohol percent by volume") + ylab('Count') + 
  ggtitle('Visualize the alcohol distribution')
```
<br>
The alcohol distribution is right skewed, log10 plot follows.
<br>
<br>

Red Wine log10 alcohol analysis.  
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_alcoholLOG, comment = NA, eval=TRUE}
ggplot(aes(alcohol), data=wineQualityReds) + 
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.02) + 
  coord_cartesian(xlim = c(8, 15)) + 
  scale_x_log10(breaks=seq(8.5, 14, 0.5)) + 
  labs(x = 'Alcohol (g / dm^3)', y = 'Count',
       title='Visualize the log10(alcohol) distribution') 
```
<br>
The alcohol log10 plot approximates a normal distribution.
<br>
<br>

Red Wine pH analysis.
```{r echo=FALSE, warning=FALSE,message=FALSE, Echo_dataset_ph, comment = NA, eval=TRUE}
summary(wineQualityReds$pH)
ggplot(data=wineQualityReds, aes(x=pH)) + 
  geom_bar(color = 'black', fill = '#ADD8E6', binwidth = 0.08) +
labs(title = 'Visualize the pH distribution', x = 'pH −log[H+]', y = 'Count') 
``` 
<br>
The pH attribute approximates a normal distribution.
<br>
<br>

Red Wine fixed acidity analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_pfixed_acidity, comment = NA, eval=TRUE}
summary(wineQualityReds$fixed.acidity)

ggplot(aes(x=wineQualityReds$fixed.acidity), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.5) +
  xlab('Fixed acidity (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the fixed acidity distribution')
```
<br>
The fixed acidity attribute approximates a normal distribution.
<br>
<br>

Red Wine volatile acidity analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_volatile_acidity, comment = NA, eval=TRUE}
summary(wineQualityReds$volatile.acidity)

ggplot(aes(x=volatile.acidity), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.08) +
  xlab('Volatile Acidity (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the volatile acidity distribution')
```
<br>
The volatile acidity attribute approximates a normal distribution.
<br>
<br>

Red Wine citric.acid analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_citric_acid, comment = NA, eval=TRUE}
summary(wineQualityReds$citric.acid)
ggplot(aes(x=wineQualityReds$citric.acid), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.03) +
  xlab('Citric Acid (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the citric acid distribution')
```
<br>
<br>

Red Wine residual sugar analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_residual_sugarSkew, comment = NA, eval=TRUE}
summary(wineQualityReds$residual.sugar)

ggplot(aes(residual.sugar), data=wineQualityReds) + 
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.5) + 
  coord_cartesian(xlim = c(0.6, 20)) + 
  scale_x_continuous(breaks=seq(0, 20, 2)) +
  labs(x = 'Residual Sugar (g / dm^3)', 
       y = 'Count', 
       title='Visualize the residual.sugar distribution')
```
<br>
The residual sugar distribution is right skewed, log10 plot follows.
<br>
<br>

Red Wine log10 residual sugar analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_residual_sugarLOG, comment = NA, eval=TRUE}
ggplot(aes(residual.sugar), data=wineQualityReds) + 
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.05) + 
  coord_cartesian(xlim = c(0.6, 20)) + 
  scale_x_log10(breaks=seq(0, 20, 2)) +
  labs(x = 'Residual Sugar (g / dm^3)', 
       y = 'Count', 
       title='Visualize the log10(residual.sugar) distribution')
```
<br>
The residual sugar log10 plot approximates a normal distribution.
<br>
<br>

Red Wine density analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_density, comment = NA, eval=TRUE}
summary(wineQualityReds$density)

ggplot(aes(x=density), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.0005) +
  xlab('Density (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the density distribution')
```
<br>
The density attribute approximates a normal distribution.
<br>
<br>

Red Wine chlorides analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_chlorides, comment = NA, eval=TRUE}
summary(wineQualityReds$chlorides)

ggplot(aes(x=chlorides), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.03) +
  xlab('Chlorides (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the chlorides distribution')
```
<br>
The chlorides distribution is right skewed, log10 plot follows.
<br>
<br>

Red Wine log10 chlorides analysis.  
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_chloridesLOG, comment = NA, eval=TRUE}
ggplot(aes(x=chlorides), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.02) +   
  coord_cartesian(xlim = c(0.04, 0.28) ) + 
  scale_x_log10(breaks=seq(0.05, 0.27, 0.04)) +
  xlab('Chlorides (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the log10(chlorides) distribution') 

```
<br>
The chlorides log10 plot approximates a normal distribution.<br>
<br>

Red Wine free sulfur dioxide analysis.
```{r echo=FALSE, warning=TRUE, message=FALSE, Echo_dataset_freeSulphurDioxide, comment = NA, eval=TRUE}
summary(wineQualityReds$free.sulfur.dioxide)

ggplot(aes(x=free.sulfur.dioxide), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin') +
  xlab('Free sulfur dioxide (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the free sulfur dioxide distribution')
```
<br>
The free sulfur dioxide distribution is right skewed, log10 plot follows. 
<br>
<br>

Red Wine log10 free sulfur dioxide analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_freeSulphurDioxideLOG, comment = NA, eval=TRUE}
ggplot(aes(free.sulfur.dioxide), data=wineQualityReds) + 
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.09) + 
  scale_x_log10(breaks=seq(10, 72, 10)) + coord_cartesian(xlim = c(3, 90)) + 
  ylab('Count') + xlab('Free sulfur dioxide (g / dm^3)') +
  ggtitle('Visualize the log10(free sulfur dioxide) distribution')
```
<br>
<br>

Red Wine total sulfur dioxide analysis.
```{r echo=FALSE, warning=TRUE, message=FALSE, Echo_dataset_total_sulfur_dioxide, comment = NA, eval=TRUE}
summary(wineQualityReds$total.sulfur.dioxide)

ggplot(aes(total.sulfur.dioxide), data=wineQualityReds) +
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin') +
  xlab('Total sulfur dioxide (g / dm^3)') + ylab('Count') + 
  ggtitle('Visualize the total sulfur dioxide distribution')
```
<br>
The total sulfur dioxide distribution is right skewed, log10 plot follows.
<br>
<br>

Red Wine log10 total sulfur dioxide analysis.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_totalSulphurDioxideLOG, comment = NA, eval=TRUE}
ggplot(aes(total.sulfur.dioxide), data=wineQualityReds) + 
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = .11) + 
  scale_x_log10(breaks=seq(0, 200, 50 )) + coord_cartesian(xlim = c(5, 200)) + 
  ylab('Count') + xlab('Total sulfur dioxide (g / dm^3)') +
  ggtitle('Visualize the log10(total sulfur dioxide) distribution')
```
<br>
The total sulfur dioxide log10 plot approximates a normal distribution.
<br>
<br>

Red Wine sulfates analysis.  
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_sulphates, comment = NA, eval=TRUE}
ggplot(aes(sulphates), data=wineQualityReds) + 
  geom_histogram(color = 'black', fill = '#ADD8E6', stat = 'bin', 
                 binwidth = 0.08) + 
  ylab('Count') + xlab('Sulphates (g / dm^3)') +
  ggtitle('Visualize the sulphates distribution') + 
  coord_cartesian(xlim = c(0.3, 1.4))
```
<br>
The sulfates attribute approximates a normal distribution.
<br>
<br>

# Univariate Analysis

### What is the structure of your dataset?
The dataset sourced from wineQualityReds.csv has 1599 entries each with 
13 features.  
  
The 11 **num** features of interest are:  
fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, 
free.sulfur.dioxide, 
total.sulfur.dioxide, density, pH, sulfates, alcohol  
  
quality is a type **int** feature  
The quality feature range is 3 through 8.  
3 represents a lower quality wine while 8 indicates a higher quality wine.  
quality was analyzed as a categorical variable.   
  
wineQualityReds.csv was downloaded from  
https://s3.amazonaws.com/
udacity-hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1530252422777000  
  
### What is/are the main feature(s) of interest in your dataset?  
My interest is exploring a possible relationship between alcohol and quality.  
For example does alcohol content appear to influence the subjective quality 
rating of the wine.  
  
I am also interested in the 4 acidity metrics pH, fixed.acidity, 
volatile.acidity, and citric acid.
I am interested in how these metrics relate to each other, and how they relate 
to the quality metric.

I will use Bivariate Plots, Bivariate Analysis, Multivariate Plots, and  
Multivariate Analysis to further my analysis of the alcohol, quality and 
acidity metrics.
<br>

The following red wine characteristics natively demonstrated a 
normal distribution -

* quality
* pH
* fixed.acidity
* volatile.acidity
* density
* sulfates
<br>

The following red wine characteristics natively demonstrated a 
skewed distribution -

* alcohol 
* residual.sugar 
* chlorides
* free.sulfur.dioxide 
* total sulfur dioxide
<br>

Follow up log10 plotting showed a near normal distribution for the following -  

* alcohol
* residual.sugar 
* chlorides
* total sulfur dioxide
<br>

Initial plot of citric.acid did not reveal a recognizable distribution.  
Follow up free.sulfur.dioxide log10 plot did not reveal a 
recognizable distribution.
<br>

# Bivariate Plots Section
```{r echo=FALSE, fig.width=10, fig.height=10,  message=FALSE, warning=FALSE, mikeTest}
library(GGally)
wineQualityReds = read.csv("wineQualityReds.csv")
wineQualityReds$quality_OF <- factor(wineQualityReds$quality, ordered = T)
dataForCorrelationStudies <- subset(wineQualityReds, select = -c(quality_OF, X))
# head(dataForCorrelationStudies,1)
ggcorr(dataForCorrelationStudies, method=c('all.obs', 'spearman'),
      nbreaks = 4, palette ='PuOr', label = TRUE,
      name='spearman correlation coeff.(rho)',
      hjust=0.8, angle=-70,size=3)  +
  ggtitle('Comprehensive Spearman Correlation Coefficient Matrix')
```
Spearman correlation interpretation.

* 0.00 - 0.19 - very weak
* 0.20 - 0.39 - weak
* 0.40 - 0.59 - moderate
* 0.60 - 0.79 - strong
* 0.80 - 1.0 - very strong
* < 0 - negative correlation
* \> 0 - positive correlation
<br>
<br>

####Focus on the relationship between quality and alcohol.  
```{r echo=FALSE, message=FALSE, quality_alcohol}
wineShort <- subset(wineQualityReds,
      select = c(fixed.acidity,volatile.acidity,citric.acid,pH,alcohol,quality))
ggpairs(wineShort[, c('quality', 'alcohol')],
        title = 'Scatterplot Matrix - Correlation Coefficient')
```  
<br>  
summary statistics - quality
```{r echo=FALSE, message=FALSE, quality_statistics}
summary(wineQualityReds$quality)
```
summary statistics - alcohol
```{r echo=FALSE, message=FALSE, alcohol}
summary(wineQualityReds$alcohol)
```
<br>

```{r echo=FALSE, message=FALSE, quality_alcohol_pH_fixed_volatile_volatile1, eval=TRUE}
ggpairs(wineShort[, c('quality', 'alcohol')],
        upper =list(continuous=wrap("smooth", color="darkgreen")),
        diag=list(continuous=wrap("barDiag", fill="red")),
        lower=list(continuous=wrap("smooth", color="blue")))
```
<br>
<br>

#### Focus on the relationship between pH and acidity metrics.  
```{r echo=FALSE, message=FALSE, quality_alcohol_citric}
ggpairs(
  wineShort[, c('pH', "volatile.acidity", "fixed.acidity", 'citric.acid')], 
  title = 'Scatterplot Matrix - Correlation Coefficients') 
```
<br>
<br>
pH summary statistics
```{r echo=FALSE, message=FALSE, pH_statistics}
summary(wineQualityReds$pH)
```
volatile acidity summary statistics
```{r echo=FALSE, message=FALSE, volatile.acidity_statistics}
summary(wineQualityReds$volatile.acidity)
```  
fixed acidity summary statistics
```{r echo=FALSE, message=FALSE, fixed.acidity_statistics}
summary(wineQualityReds$fixed.acidity)
```  
citric acid summary statistics
```{r echo=FALSE, message=FALSE, citric.acidity_statistics}
summary(wineQualityReds$citric.acid)
```
<br>
 
```{r echo=FALSE, message=FALSE, quality_alcohol_pH_citric1}
ggpairs(wineShort[, c('pH', "volatile.acidity", "fixed.acidity", 
                      'citric.acid')],
        upper =list(continuous=wrap("smooth", color="darkgreen")),
        diag=list(continuous=wrap("barDiag", fill="red")),
        lower=list(continuous=wrap("smooth", color="blue")))
```
<br>
<br>

Red Wine alcohol - quality analysis, alcohol percent by volume.
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_dataset_alcohol_quality, comment = NA, eval=TRUE}
wineQualityReds$quality_OF <- factor(wineQualityReds$quality, ordered = T)
ggplot(wineQualityReds) + geom_boxplot(aes(x = quality_OF, y = alcohol),
                                       color =  '#3299CC') +
  xlab('Quality rating') + ylab('Alcohol percent by volume')
```
<br>
Higher quality rated wines, 6, 7, and 8 contain progressively higher alcohol.
<br>
<br>

Red Wine alcohol - quality analysis, quality percent by volume.
```{r echo=FALSE, warning=FALSE, message=TRUE, Echo_dataset_alcohol_quality_scatter1, comment = NA, eval=TRUE}
m1 <- ggplot(aes(x = quality, y = alcohol), data = wineQualityReds) +
    geom_point(position = position_jitter(h = 50)) + 
    xlim(2.9, quantile(wineQualityReds$quality, 0.95)) +
    ylim(8, quantile(wineQualityReds$alcohol, 0.95)) + 
    geom_smooth(method = 'lm', color = 'red') + 
    xlab('Quality rating') + ylab('Alcohol percent by volume') +
    labs(caption='Data Source: Udacity Project Data Set Options')
# title = 'Red Wine alcohol - quality analysis') 
m1
```
<br>
Quantify the strength of the quality - alcohol relationship.  
Pearson’s r
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_alcohol_quality_correlation, comment = NA, eval=TRUE}
# cor.test(wineQualityReds$quality, wineQualityReds$alcohol, method = 'pearson')
with(wineQualityReds, cor.test(quality, alcohol, method = 'pearson'))
```  
Correlation Interpretation  
quality - alcohol correlation is greater than the accepted meaningful 
correlation threshold - 0.3  
quality - alcohol correlation less than the accepted moderate correlation 
threshold - 0.5  

Leverage the above quality - alcohol visualizations to fine tune correlation 
analysis.  
Quantify the strength of the quality - alcohol relationship - quality > 4.  
Pearson’s r
```{r echo=FALSE, warning=TRUE, message=TRUE, Echo_alcohol_quality_correlation_subset, comment = NA, eval=TRUE}
with(subset(wineQualityReds, quality > 4), 
     cor.test(quality, alcohol, method = 'pearson'))
```  
Correlation Interpretation - subset quality > 4  
subset quality - alcohol correlation is greater than the accepted moderate 
correlation threshold - 0.5  
subset quality - alcohol correlation is less than the accepted large, strong 
correlation threshold - 0.7  
Higher quality red wines have a higher alcohol correlation.
<br>
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_alcohol_analysis_by_quality, eval=TRUE}
redsGroupedByQuality <- group_by(wineQualityReds, quality)
red_alcohol_mean_quality_grouping <- summarise(redsGroupedByQuality, 
                                               average = mean(alcohol), 
                                               count=n())
ggplot(aes(x = quality, y = average ), 
       data = red_alcohol_mean_quality_grouping) +
  geom_line(stat='identity', colour="#000099") + ylab('average EtOH') + 
  ggtitle('mean (average) EtOH by quality group') + 
  scale_y_continuous(breaks = seq(9,13,1)) + 
  coord_cartesian(ylim = c(9,13)) + 
  xlab('Quality rating') + ylab('Average alcohol percent by volume')
```
<br>
Alcohol and quality proportional.
<br>
<br>

#### quality - acidity metrics analysis
#### sorted by Pearson's r correlation ascending
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_volatile_acidity_quality, eval=TRUE}
ggplot(wineQualityReds) + geom_boxplot(aes(x = quality_OF, 
                                           y = volatile.acidity)) +
   xlab('Quality rating') +  ylab('Volatile Acidity - (g / dm^3)') + 
  ggtitle('Red wine quality - volatile acidity')
```

Pearson's r - quality - volatile acidity correlation 
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_volatile_acidity_quality_pearson, eval=TRUE}
cor.test(wineQualityReds$quality, wineQualityReds$volatile.acidity, 
         method = 'pearson') 
```
quality - volatile acidity negative correlation - meaningful but weak
<br>
<br>


```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_pH_quality, eval=TRUE}
ggplot(wineQualityReds) + geom_boxplot(aes(x = quality_OF, y = pH)) + 
  labs(title = 'Red wine quality - pH', x = 'Quality rating', y = 'pH −log[H+]')
```

Pearson's r - quality - pH correlation 
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_pH_quality_pearson}
cor.test(wineQualityReds$quality, wineQualityReds$pH, method = 'pearson')  
```
quality - pH negative correlation less than the accepted weak threshold - -0.3
<br>
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_fixed_acidity_quality, eval=TRUE}
ggplot(wineQualityReds) + geom_boxplot(aes(x = quality_OF, y = fixed.acidity)) + 
  labs(title = 'Red wine quality - fixed acidity', 
       x = 'Quality rating', y = 'Fixed Acidity - (g / dm^3)')
```

Pearson's r - quality - fixed acidity correlation 
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_fixed_acidity_quality_pearson}
cor.test(wineQualityReds$quality, wineQualityReds$fixed.acidity, 
         method = 'pearson') # 0.1240516 
```
quality - fixed acidity correlation less than the accepted weak threshold - 0.3
<br>
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_citric_acid_quality, eval=TRUE}
ggplot(wineQualityReds) + geom_boxplot(aes(x = quality_OF, y = citric.acid)) + 
  labs(title = 'Red wine quality - citric acid', x = 'Quality rating', 
       y = 'Citric Acid - (g / dm^3)')
```

Pearson's r - quality - citric.acid correlation 
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_citriv_acid_quality_pearson}
cor.test(wineQualityReds$quality, wineQualityReds$citric.acid, 
         method = 'pearson') 
```
quality - citric acid correlation less than the accepted weak threshold - 0.3
<br>
<br>

#### acidity metrics - pH analysis
#### sorted by Pearson's r correlation ascending
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_fixed_acidity_pH}
ggplot(data = wineQualityReds, aes(x = pH, y = fixed.acidity)) +
  geom_point(alpha = 1/5) + xlab('pH −log[H+]') + 
  ylab('Fixed acidity - (g / dm^3)') +
  geom_smooth(method="lm") + ggtitle('Red wine pH - fixed acidity') 
```

Pearson's r - pH - fixed acidity correlation 
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_fixed_acidity_pH_pearson}
cor.test(wineQualityReds$pH, wineQualityReds$fixed.acidity, 
         method = 'pearson') 
```
pH - fixed acidity - moderate correlation
<br>
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_citric_acid_pH, eval=TRUE}

ggplot(data = wineQualityReds, aes(x = pH, y = citric.acid)) +
  geom_point(alpha = 1/5) + ylab('citric acid') +
  geom_smooth(method="lm") + 
  labs(title = 'Red wine pH - citric acid',  x = 'pH −log[H+]', 
       y = 'Citric Acid')
```

Pearson's r - pH - citric acid correlation 
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_citric_acid_pH_pearson}
cor.test(wineQualityReds$pH, wineQualityReds$citric.acid, 
         method = 'pearson') 
```
pH - citric acid - moderate correlation  
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_volatile_acidity_pH, eval=TRUE}
ggplot(data = wineQualityReds, aes(x = pH, y = volatile.acidity)) +
  geom_point(alpha = 1/5) + ylab('volatile acidity') +
  geom_smooth(method="lm") + scale_y_continuous(breaks = seq(0.12, 1.22, 0.2)) +
  coord_cartesian(ylim = c(0.12,1.22)) + 
  labs(title = 'Red wine pH - volatile acidity',  x = 'pH −log[H+]', y = 'Volatile Acidity')
```

Pearson's r - pH - volatile acidity correlation 
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Echo_volatile_acidity_pH_pearson}
cor.test(wineQualityReds$pH, wineQualityReds$volatile.acidity, 
         method = 'pearson') 
```
pH - volatile acidity correlation less than the accepted weak threshold - 0.3
<br>
  
# Bivariate Analysis  
##### Red wine: quality rating - alcohol content analysis.  
* The red wine quality rating and alcohol content exhibited a positive 
correlation.
* The higher the red wine quality the stronger the quality - 
alcohol correlation.
* The one exception is the drop in alcohol content associated with quality 
rating increase from 4 to 5.

##### Next the relationships between the quality ratings 
##### and the acidity metrics were analyzed.  
* red wine quality - volatile acidity
* red wine quality - pH
* red wine quality - fixed acidity
* red wine quality - citric acid

* summary: quality - acidity metrics observations  
    + volatile acidity had the strongest negative correlation - meaningful but 
    weak
    + citric acid had the strongest correlation - although that correlation is 
    less than weak
    
##### Next the relationships between the acidity metrics and pH were analyzed.
reminder - pH and acidity are inversely proportional 

* pH - fixed acidity - strongest correlation - strength: moderate
* pH - citric acid - correlation strength: moderate  
* pH - volatile acidity - correlation strength: less than weak  
  
# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, pHAlcoholQuality}
 ggplot(data = wineQualityReds, aes(x = alcohol, y = pH)) + 
  facet_wrap(~ quality_OF)  + geom_smooth(method = 'loess') + 
  geom_jitter(alpha=.2) + 
  stat_cor(method = 'pearson', label.x.npc = 0.6, label.y.npc = 0.08)  + 
  labs(title = "pH - alcohol by quality - Pearson's r", x = 'Alcohol percent by volume', 
       y = 'pH −log[H+] - lower value - more acidic')
```  
<br>
For most quality ratings, alcohol increase is accompanied by pH 
increase (lower acidity).
<br>
<br>

Red wine pH, quality follow up analysis.  
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, pHAlcohol}
p1 <- ggplot(wineQualityReds, 
             aes(as.factor(signif(wineQualityReds$pH,2)), quality)) + 
  geom_boxplot() + coord_cartesian(xlim = c(2,6)) + 
  labs(title = 'Visualize pH - quality', 
     x = 'pH −log[H+]', y = 'Quality rating')
  
p2 <- ggplot(wineQualityReds, aes(as.factor(floor(alcohol)), quality)) + 
  geom_boxplot() + labs(title = 'Visualize Alcohol - quality', 
     x = 'Alcohol percent by volume', y = 'Quality rating')

grid.arrange(p1, p2)
```
<br>
Quality tends to increase as pH, alkalinity increases.  
Quality tends to increase as alcohol percent by volume increases.
<br>
<br>

```{r Alcohol vs. Quality and pH, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
 
tempWineQualityReds <- read.csv("wineQualityReds.csv")
tempWineQualityReds <- filter(tempWineQualityReds, !is.na(pH))
tempWineQualityReds <- subset( tempWineQualityReds, pH > 3 & pH < 3.9)

ggplot(aes(x = alcohol, y =  quality,
           color =  droplevels(factor(signif(pH, 2)), exclude = c(2.7, 4))), 
           data = tempWineQualityReds) +
  geom_point(size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'seq', palette = "BuGn",
                     guide = guide_legend(title = 'pH', reverse = F, 
                                          override.aes = list(size = 2))) + 
  coord_cartesian(xlim=c(8,15)) + 
  labs(title = 'alcohol - quality - pH', x = 'Alcohol percent by volume', y = 'Quality rating') 
```
<br>
A comprehensive definitive pH trend is not illustrated.  
No discernible color gradient pattern, striations observed.
<br>
<br>

```{r AlcoholQualityFixedAcidity, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y =  quality,
           color = factor(signif(fixed.acidity, 1))), 
           data = wineQualityReds) +
  geom_point(size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'seq',  palette = "GnBu",
                     guide = guide_legend(title = 'Fixed Acidity', reverse = F, 
                                          override.aes = list(size = 2))) + 
  labs(title = 'alcohol - quality - fixed acidity', 
       x = 'Alcohol percent by volume', y = 'Quality rating')
```
<br>
A comprehensive definitive Fixed Acidity trend is not illustrated.  
No discernible color gradient pattern, striations observed.
<br>
<br>
 
```{r AlcoholQualityVolatileAcidity, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
tempWineQualityReds <- read.csv("wineQualityReds.csv")
tempWineQualityReds <- filter(tempWineQualityReds, !is.na(volatile.acidity))
tempWineQualityReds$volatile.acidity <- 
  signif(tempWineQualityReds$volatile.acidity, 1)
tempWineQualityReds <- 
  subset( tempWineQualityReds, volatile.acidity > 0.1 & volatile.acidity < 2)

ggplot(aes(x = alcohol, y =  quality,
           color = factor(signif(volatile.acidity, 1))), 
           data = tempWineQualityReds) +
  geom_point(size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'seq', palette = "Blues",
                     guide = guide_legend(title = 'Volatile Acidity', 
                     reverse = F, override.aes = list(size = 2))) + 
  labs(title = 'alcohol - quality - volatile acidity', 
       x = 'Alcohol percent by volume', y = 'Quality rating' )
```
<br>
A comprehensive definitive Volatile Acidity trend is not illustrated.  
No discernible color gradient pattern, striations observed.
<br>
<br>

```{r AlcoholQualityCitricAcid, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y =  quality, 
           color = cut(citric.acid, include.lowest = TRUE, 
                       breaks = seq(0, 1, by = .1))), 
       data = tempWineQualityReds) + geom_point(size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'seq', palette = "Reds", 
                     guide = guide_legend(title = 'Citric Acid', 
                     reverse = F, override.aes = list(size = 2))) + 
  labs(title = 'alcohol - quality - citric acid', 
       x = 'Alcohol percent by volume', y = 'Quality rating')
```
<br>
A comprehensive definitive Citric Acid trend is not illustrated.  
No discernible color gradient pattern, striations observed.
<br>
<br>

# Multivariate Analysis

For  most quality ratings, alcohol increase is accompanied by pH increase 
(lower acidity).  
<br>
There was not a comprehensive explicit relationship seen when pH was color  
plotted on the quality alcohol plot.  
<br>
There was not a comprehensive explicit relationship seen when Fixed Acidity was  
color plotted on the quality alcohol plot.  
<br>
There was not a comprehensive explicit relationship seen when Volatile Acidity  
was color plotted on the quality alcohol plot.  
<br>
There was not a comprehensive explicit relationship seen when Citric Acid was  
color plotted on the quality alcohol plot.  

------  

# Final Plots and Summary

<!-- ### comment syntax example -->
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Plot_One}
ggplot(data = wineQualityReds, aes(x = alcohol, y = pH)) + 
  facet_wrap(~ quality_OF)  + geom_smooth(method = 'loess') + 
  geom_jitter(alpha=.2) + 
  stat_cor(method = 'pearson', label.x.npc = 0.6, label.y.npc = 0.08)  + 
  labs(title = "pH - alcohol by quality - Pearson's r", x = 'alcohol content', 
       y = 'pH −log[H+] - lower value - more acidic')
# knitr::knit_exit()
```  
<br>
For most quality ratings, alcohol increase is accompanied by pH increase 
(lower acidity).  
I chose this plot because it communicates the relationship between alcohol 
content and pH,  
broken out by quality rating.  
Pearson's r mathematically augments the visualizations.
<br>
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, Plot_Two}

redsGroupedByQuality <- group_by(wineQualityReds, quality)
red_alcohol_mean_quality_grouping <- 
  summarise(redsGroupedByQuality, average = mean(alcohol), count=n())
# red_alcohol_mean_quality_grouping
ggplot(aes(x = quality, y = average ), 
       data = red_alcohol_mean_quality_grouping) +
  geom_line(stat='identity', colour="#000099") + 
  ylab('Alcohol percent by volume') + xlab('Quality rating') + 
  ggtitle('mean (average) EtOH by quality group') + 
  scale_y_continuous(breaks = seq(9,13,1)) + coord_cartesian(ylim = c(9,13)) 
```
<br>
I chose this plot because it communicates the strong relationship between 
alcohol content and quality rating.  
Per the Centers for Disease Control and Prevention -  
"Alcohol use slows reaction time and impairs judgment..."      
CDC's acknowledgement of the effect alcohol has on judgement indicates further 
analysis is indicated.  
Does alcohol content effect subjective quality rating reporting?  
Further analysis would include a controlled experiment where alcohol 
concentration was the only variable.  
The results would be used to further analyze and refine the relationship 
between alcohol concentration  
and the subjective quality rating.   
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, plot_Three}
ggplot(data = wineQualityReds, aes(x = pH, y = volatile.acidity)) +
  geom_point(alpha = 1/5) + xlab('pH −log[H+]') + 
  ylab('Volatile Acidity - (g / dm^3)') +
  geom_smooth(method="lm") + scale_y_continuous(breaks = seq(0.12, 1.22, 0.2)) +
  coord_cartesian(ylim = c(0.12,1.22)) + 
  ggtitle('Red wine pH - volatile acidity')
```
<br>
I chose this plot because of the observed relationship between 
Volatile Acidity and pH.  
The plot suggests a rise in pH (more alkaline, less acid measurements), as 
Volatile Acidity increases.  
Initially one might assume if a measure of acidity such as Volatile Acidity 
increased  
there would be a corresponding decrease in pH, a more acidic pH measurement.
<br>
<br>

```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, pHAlcohol_quality}
tempWineQualityReds <- read.csv("wineQualityReds.csv")
tempWineQualityReds <- na.omit(tempWineQualityReds)

tempWineQualityReds$quality.levels <- 
  cut(tempWineQualityReds$quality, breaks = c(2, 4, 6, 8), right = T, labels = c('LOW.Q', 'MIDDLE.Q','HIGH.Q')) 

ggplot(data = tempWineQualityReds, aes(pH, alcohol, 
                                       color = quality.levels)) + 
  coord_cartesian(xlim = c(2.7, 4), ylim = c(7, 14)) +
  geom_jitter(size=1) + 
  geom_smooth(method=lm) + 
   scale_x_continuous(breaks = seq (2.7, 4, 0.2)) +
   scale_y_continuous(breaks = seq (7, 14, 0.4)) + 
   scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality levels', reverse = F, 
                                          override.aes = list(size = 2))) +
  theme_dark() + 
  labs(title = 
         'Relationship of pH and alcohol with colored quality levels', 
       x = 'pH −log[H+]', y = 'Alcohol percent by volume')
```
<br>
I chose this plot because it illustrates the relationship between:  
  
* alcohol percent by volume  
* pH  
* quality levels  

Higher quality wines tend to have centric pH values, and higher alcohol content.
<br>
<br>

Additional studies - density - alcohol - quality analysis.
```{r echo=FALSE, warning=FALSE, message=FALSE, comment = NA, densityAlcohol_quality}
 ggplot(data = tempWineQualityReds, aes(density, alcohol, 
                                       color = quality.levels)) + 
  coord_cartesian(xlim = c(0.989, 1.005), ylim = c(7,5, 15)) +
  geom_jitter(size=1) + 
  geom_smooth(method=lm) + 
   scale_x_continuous(breaks = seq (0.985, 1.005, 0.002)) +
   scale_y_continuous(breaks = seq (7.5, 15, 2.5)) + 
   scale_color_brewer(type = 'seq',
                     guide = guide_legend(title = 'Quality levels', reverse = F, 
                                          override.aes = list(size = 2))) +
  theme_dark() + 
  labs(title = 
         'Relationship of density and alcohol with colored quality levels', 
       x = 'Density (g / dm^3)' , y = 'Alcohol percent by volume') 
```
<br>
I chose this plot because it illustrates the relationship between:  
  
* alcohol percent by volume  
* density  
* quality levels  

Higher quality wines tend to have lower density and higher alcohol content,
<br>

------

# Reflection

What were some of the struggles you went through?  
<ul>
<li>Data preparation for plotting was tedious and time consuming.</li>
<li>Creating the needed ordered factor variable(s) 
from the original variables.</li>
<li>Optimizing the scale and increment of the x and y axis to optimally 
communicate the underlying data.</li>
<li>Configure function arguments such as alpha and jitter to facilitate the 
rendering of the plotted data.</li>
<li>Manage levels of factored data sets, and palettes to emphasize data trends, 
while minimizing outliers.</li>
</ul>  

What was surprising?  
The unexpected relationship between Volatile Acidity and pH.  
As Volatile Acidity increased, pH increased.  pH is a measure of acidity.  
It was unexpected to see the Volatile Acidity increase yield a more alkaline, 
less acidic pH result.  

Future work.  
Further investigate the relationship between alcohol content and subjective 
quality rating.  
Next step - Additional quality rating analysis based on a 
controlled experiment where alcohol concentration  
is the only variable.  
<br>
<br> 
